---
title: "the awesomest project ever true?"
author: "Morde Bernstein, Marley McGehee, ________"
date: "3/25/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

## Introduction

## Data Import

### Loading the Data

```{r}
coral <- read.csv("Global_Coral_Bleaching_Database.csv",na="N/A")
coral <- janitor::clean_names(coral)
coral
```

```{r}
coral_relevant_columns <- coral%>%
  summarize(
    record_id=record_id,
    percent_mortality=percent_mortality,
    coral_regions=coral_regions,
    max_percent_bleached=max_percent_bleached
    )
coral_relevant_columns
```
### Cleaning Perecent Mortality

```{r, warning = FALSE, message = FALSE}
columns_coral_relevant_columns <- coral_relevant_columns %>%
  separate_wider_delim(percent_mortality, delim = "-",
names = c("mortality_one", "mortality_two"),
too_few = "align_start",
too_many = "merge"
) %>%
  mutate(
    mortality_one = parse_number(mortality_one),
    mortality_two = parse_number(mortality_two) )

columns_coral_relevant_columns
```

```{r}
means_coral <- columns_coral_relevant_columns %>%
  mutate(average_percent_mortality = rowMeans(select(.,mortality_one, mortality_two), na.rm = TRUE)) %>%
  filter(average_percent_mortality < 100) %>%
  mutate(max_percent_bleached = parse_number(max_percent_bleached))


means_coral
```




Some of the data for percent mortality were given in ranges, and some of it was single values. In order to use this variable, we separated the data into two separate columns, mortality one and mortality two. Mortality one had the values that were the lower end of the range as well as the single values, and mortality two had the higher end of the ranges. We then cleaned the data and removed any words or units that were included. In order to have a single variable, we found the means for mortality one and two by using row means. Following this, we filtered by less than 100 to get rid of any values that were leftover years from the descriptions.  Additionally, we ran max percent bleached through parse number to make sure it was numeric and could be used for calculations. 


## Univariate Data


## Bivariate Data

### Max Percent Bleached vs Coral Regions



### Max Percent Bleached vs Percent Mortality

#### Scatterplot with Smoothed Curve

```{r}
ggplot(means_coral, aes(x = max_percent_bleached, y =average_percent_mortality )) +
  geom_point(alpha = 0.2)+ 
  geom_smooth(color = "blue")
```

#### Correlation

```{r}
knitr::kable(correlation_coral<- means_coral%>% filter(!is.na(max_percent_bleached))%>%
  summarize(correlation = cor(max_percent_bleached,average_percent_mortality)))
```



There appears to be many average percent mortality rates of zero, which is weighing down the trend of the data. Additionally there seems to be many data points clustered towards 0 and 100 percent max bleached. Logically, it makes sense that as the percent bleached increases so does the percent mortality, but that is hard to see if that is true from the data due to a low correlation of 0.22. 


## Trivariate Data

## Conclusion







