---
title: "the awesomest project ever true"
author: "Morde Bernstein, Marley McGehee, Kimberly McNeil"
date: "3/25/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

## Introduction

## Data Import

### Loading the Data

```{r}
coral <- read.csv("Global_Coral_Bleaching_Database.csv",na="N/A")
coral <- janitor::clean_names(coral)
coral
```

```{r}
coral_relevant_columns <- coral%>%
  summarize(
    record_id=record_id,
    percent_mortality=percent_mortality,
    coral_regions=coral_regions,
    max_percent_bleached=max_percent_bleached
    )
coral_relevant_columns
```
### Cleaning Perecent Mortality

```{r, warning = FALSE, message = FALSE}
columns_coral_relevant_columns <- coral_relevant_columns %>%
  separate_wider_delim(percent_mortality, delim = "-",
names = c("mortality_one", "mortality_two"),
too_few = "align_start",
too_many = "merge"
) %>%
  mutate(
    mortality_one = parse_number(mortality_one),
    mortality_two = parse_number(mortality_two) )

columns_coral_relevant_columns
```

```{r}
means_coral <- columns_coral_relevant_columns %>%
  mutate(average_percent_mortality = rowMeans(select(.,mortality_one, mortality_two), na.rm = TRUE)) %>%
  filter(average_percent_mortality < 100) %>%
  mutate(max_percent_bleached = parse_number(max_percent_bleached))


means_coral
```




First we loaded the data and cleaned the names. Following that we saved only our target variables that we would be looking at. Some of the data for percent mortality were given in ranges, and some of it was single values. In order to use this variable, we separated the data into two separate columns, mortality one and mortality two. Mortality one had the values that were the lower end of the range as well as the single values, and mortality two had the higher end of the ranges. We then cleaned the data and removed any words or units that were included. In order to have a single variable, we found the means for mortality one and two by using row means. Following this, we filtered by less than 100 to get rid of any values that were leftover years from the descriptions.  Additionally, we ran max percent bleached through parse number to make sure it was numeric and could be used for calculations. 


## Univariate Data

### Coral Regions

[description of what this variable is]

#### Visual Summary

```{r}
means_coral %>%
  count(coral_regions) %>%
  ggplot(aes(x = reorder(coral_regions, n), y = n)) +
  geom_col(fill = "steelblue",
           color = "black") +
  coord_flip() +
  labs(title = "Bar Plot of Categories",
       x = "Category",
       y = "Count")
```


#### Numeric Summary

```{r}
means_coral %>%
  count(coral_regions) %>%
  mutate(proportion_round = round((n / sum(n)) / 0.0001) * 0.0001) %>%
  kable(caption = "Frequency Table for Coral Regions",
        col.names = c("Coral Regions", 
                      "n", 
                      "Proportion"))
```





### Maximum Percent Bleached

```{r}
means_coral %>% 
  ggplot(aes(x = max_percent_bleached)) +
    geom_histogram(bins = 15,
                   fill = "steelblue",
                   color = "black") +
    labs(title = "Distribution of Maximum Percent Bleached (Histogram)",
         x = "Value",
         y = "Count")
```

```{r}
ggplot(means_coral, aes(x = max_percent_bleached)) +
  geom_density(fill = "steelblue",
               color = "black") +
    labs(title = "Distribution of Maximum Percent Bleached (Density)",
         x = "Value",
         y = "Density")
```



#### Numeric Summary

```{r}
means_coral %>%
  summarise(
    Count = n(),
    Mean = mean(max_percent_bleached, na.rm = TRUE),
    Median = median(max_percent_bleached, na.rm = TRUE),
    sd = sd(max_percent_bleached, na.rm = TRUE),
    Min = min(max_percent_bleached, na.rm = TRUE),
    Q1 = quantile(max_percent_bleached, 0.25, na.rm = TRUE),
    Q3 = quantile(max_percent_bleached, 0.75, na.rm = TRUE),
    Max = max(max_percent_bleached, na.rm = TRUE)
  ) %>% 
  kable(caption = "Summary Statistics for Maximum Percent Bleached")
```



### Average Percent Mortality

```{r}
means_coral %>% 
  ggplot(aes(x = average_percent_mortality)) +
    geom_histogram(bins = 15,
                   fill = "steelblue",
                   color = "black") +
    labs(title = "Distribution of Average Percent Mortality (Histogram)",
         x = "Value",
         y = "Count")
```

```{r}
ggplot(means_coral, aes(x = average_percent_mortality)) +
  geom_density(fill = "steelblue",
               color = "black") +
    labs(title = "Distribution of Average Percent Mortality (Density)",
         x = "Value",
         y = "Density")
```



#### Numeric Summary

```{r}
means_coral %>%
  summarise(
    Count = n(),
    Mean = mean(average_percent_mortality, na.rm = TRUE),
    Median = median(average_percent_mortality, na.rm = TRUE),
    sd = sd(average_percent_mortality, na.rm = TRUE),
    Min = min(average_percent_mortality, na.rm = TRUE),
    Q1 = quantile(average_percent_mortality, 0.25, na.rm = TRUE),
    Q3 = quantile(average_percent_mortality, 0.75, na.rm = TRUE),
    Max = max(average_percent_mortality, na.rm = TRUE)
  ) %>% 
  kable(caption = "Summary Statistics for Average Percent Mortality")
```



## Bivariate Data

### Max Percent Bleached vs Coral Regions

#### Visual Summary

```{r}
ggplot(means_coral, aes(x=coral_regions, y = max_percent_bleached))+
  geom_boxplot() +
  labs (x = "Coral Regions", y = "Max Percent Bleached")
```

#### Numberic Summary

```{r}
knitr::kable(means_coral %>%
  group_by(coral_regions) %>%
  summarize(
    n = n(),
    first_quantile = quantile(max_percent_bleached, 0.25, na.rm = TRUE),
    median = median(max_percent_bleached, 0.5, na.rm = TRUE),
    third_quantile = quantile(max_percent_bleached, 0.75, na.rm = TRUE),
    mean = mean(max_percent_bleached, na.rm = TRUE),
    standard_deviation = sd(max_percent_bleached, na.rm = TRUE)
 ))
```


All four of the reef types have the same range for maximum percent bleached, 0 to 100. West Indian Reef tends to have higher maximum percent bleach compared to the other three reefs. It has the highest mean maximum percent bleached. Great Barrier Reef and other reefs are similar to each other with low first quartile and median percent bleaching compared to the West Indian Reef and Caribbean/GoM reefs. However, there is quiet a large standard deviation for all four of the reef types.

### Max Percent Bleached vs Percent Mortality

#### Scatterplot with Smoothed Curve

```{r}
ggplot(means_coral, aes(x = max_percent_bleached, y =average_percent_mortality )) +
  geom_point(alpha = 0.2)+ 
  geom_smooth(color = "blue") +
  labs(x = "Max Percent Bleached", y = "Average Percent Mortality")
```

#### Correlation

```{r}
knitr::kable(correlation_coral<- means_coral%>% filter(!is.na(max_percent_bleached))%>%
  summarize(correlation = cor(max_percent_bleached,average_percent_mortality)))
```



There appears to be many average percent mortality rates of zero, which is weighing down the trend of the data. Additionally there seems to be many data points clustered towards 0 and 100 percent max bleached. Logically, it makes sense that as the percent bleached increases so does the percent mortality, but that is hard to see if that is true from the data due to a low correlation of 0.22. 


## Trivariate Data

### Average Percent Mortality vs Max Percent Bleached, separated by Region

```{r}
means_coral%>%
  ggplot(mapping=aes(max_percent_bleached,average_percent_mortality))+
  geom_point(alpha=.2)+
  geom_smooth()+
  facet_wrap(~coral_regions)
```
The scatterplots. I'll do the correlations and analysis soon

## Conclusion







